---
title: "Getting started with reemscenex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with reemscenex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.19
)
```

```{r setup}
library(reemscenex)
library(ggplot2)
```

The goal of the `{reemscenex}` package is to extend `{reem}` to be able to simulate hypothetical scenarios in a forecast. The general procedure is to start with a `{reem}` model object that has previously been fitted to data, update its parameters to reflect a new scenario, and simulate the model.

In this vignette, we use a `{reem}` fitted model object built into `{reemscenex}` called `model.example` in order to demonstrate the functionality of the package.

# A note about `{reem}` model objects

`{reem}` uses the object-oriented [Reference Class](http://adv-r.had.co.nz/R5.html) system, which has three main differences worth noting to the usual functional programming approach adopted by most R packages:

- The functions (methods) that can operate on a `{reem}` model object are bundled in the object (as opposed to called on the object from the package). For instance, `{reem}`'s forecast method is called on a `model` object like `model$forecast()` (object-oriented), as opposed to `reem::forecast(model)` (functional). 
- A `{reem}` model object's methods **modify the object in memory**! For instance, if you use the `model$forecast()` method, the results of the forecast are saved in the `model` going forward. 
- The usual assignment operator (`<-` or `=`) **does not copy the object over** for Reference classes, but instead aliases it with another name. Doing
  ```{r}
  #| eval: FALSE
  model.fitted <- model.example
  model.fitted$forecast()
  ```
  **would modify `model.example` and `model.fitted`** by adding the forecast output to them (because they are the same object)! This feature is worth bearing in mind and it's usually safer to use each model object's built-in `$copy()` method as we do now to avoid keep around a "clean" copy of an example model object as `model.example`:
  ```{r}
  model.fitted <- model.example$copy()
  ```

# Generating forecasts

We can forecast a scenario using the `reemscenex::fcst_sc()` function. This function uses the model object's built-in `$forecast()` method.

## Status quo forecast

If we just run `fcst_sc()` on the model object, it will produce a status quo forecast with the parameters and forecast settings already baked into `model.fitted` (attached to the object when the model is fitted): 

```{r, warnings = FALSE}
fcst1 <- fcst_sc(model=model.fitted)
```

This function returns a list of outputs:

```{r}
class(fcst1)
names(fcst1)
```

The documentation for `fcst_sc()` tells us what this output is:

```{r}
?fcst_sc
```

For instance, here is the forecast summary:

```{r}
head(fcst1$summary.fcst)
```

Again, because of how Reference Classes work, this output is also saved within the `model.fitted` object in memory, under the `fcst.obj` element:

```{r}
names(model.fitted$fcst.obj)

# here is the forecast summarized by daily quantiles
head(model.fitted$fcst.obj$summary.fcst)
```

For more information about forecast outputs, as well as `{reem}` itself, you can review [the `{reem}` package README](https://github.com/phac-nml-phrsd/reem/blob/main/README.md).

## Forecasting with a scenario

If we want to forecast with a scenario, _i.e._ by updating the baseline (non-fitted) parameters, we can use the `prm.sc` argument of `fcst_sc()`. (However, this is not the recommended approach, and we will present a better way to accomplish this task in the next section.) 

For instance, the default value for the hospitalization proportion, `h.prop` is:

```{r}
model.fitted$prms$h.prop
```

We can increase this by an order of magnitude to see a clear effect in the forecast:

```{r}
fcst2 <- fcst_sc(
  model=model.fitted,
  prm.sc=list(h.prop = model.fitted$prms$h.prop*10) 
)
```

```{r}
#| echo: false

# vignette functions, just for demo
prep_fcst <- function(fcst, model, vars){
  (reem:::summarize_fcst(fcst$simfwd, model$fcst.prm, vars = vars) 
   |> tidyr::pivot_wider(
  id_cols = c("name", "date", "mean"),
  names_from = "qprob",
  values_from = "q"
  ))
}

plot_fcst <- function(sim){
  pp <- ggplot(sim, aes(x = date))
  
  if("fcst" %in% colnames(sim)){
    pp <- (pp
      + geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`, fill = fcst), alpha = 0.5) 
      + geom_line(aes(y = mean, colour = fcst), linewidth = 1.25)     
    )
  } else {
    pp <- (pp
      + geom_ribbon(aes(ymin = `0.05`, ymax = `0.95`), alpha = 0.5) 
      + geom_line(aes(y = mean), linewidth = 1.25)     
    )
  }
  
  (pp + theme(axis.title = element_blank()))
}

annotate_x <- function(gg, x, y.pos, label, font.size = 3){
    y.range <- layer_scales(gg)$y$get_limits()
    label.df <- data.frame(
        x = x,
        y = y.range[1] + y.pos * (y.range[2] - y.range[1]),
        label = label
    )
    (gg
        + geom_vline(data = label.df, mapping = aes(xintercept = x), linetype="dashed", alpha = 0.5)
        + geom_label(data = label.df,  mapping = aes(x = x, y = y, label = label), alpha = 0.5, size = font.size)
    )
}
```

```{r}
# prep_fcst() and plot_fcst() are utilities defined in this vignette
# and are not included in the `{reemscenex}` package

(lapply(
  list(fcst1, fcst2), 
  \(x){prep_fcst(x, model=model.fitted, vars=c("H"))})
  |> dplyr::bind_rows(.id = "fcst") 
  |> plot_fcst()
  + labs(title = "Predicted hospitalizations")
)
```

We see above that increasing `h.prop` greatly increased the predicted hospitalizations over time.

Updating model parameters via `fcst_sc()` will change the `model.fitted` object in memory:

```{r}
model.fitted$prms$h.prop
```

## Scenario functions

Scenarios are fundamentally built by modifying the model object's parameters, though doing this manually as demonstrated above is not advised. Instead, `{reemscenex}` implements an `sc()` function that carefully sets up a pre-defined scenario in the model, validating inputs and ensures requested scenarios make sense with the underlying fitted model:

```{r}
# gradual change in transmission
# reset model object
model.fitted <- model.example$copy()

start.date <- as.Date("2022-01-30")
end.date <- as.Date("2022-03-30")
value <- 2

sim <- (
  # define scenario
    sc(
    model = model.fitted, 
    opts = list(
      name = "gradual change in transmission",
      start.date = start.date,
      end.date = end.date,
      value = value,
      epsilon = 0.001
    )
  )
  # forecast from the scenario
  |> fcst_sc(model = model.fitted)
)

(sim
  |> prep_fcst(model=model.fitted, vars=c("Y"))
  |> plot_fcst()
  |> annotate_x(
      x = start.date, y.pos = 0.8,
      label = stringr::str_wrap(glue::glue("transmission increases gradually by {value}x between {format(start.date, '%d %b')} and {format(end.date, '%d %b')}"), width = 30)
    )
  + labs(title = "Predicted clinical reports")
)
```

A full description of pre-defined scenarios can be found in `vignette("scenarios")`. You can remind yourself of the scenarios with short descriptions and required options for each in the Details section of `?sc`. 

## Composing scenarios

Scenarios can be composed by chaining successive calls to `sc()`:

```{r}
# reset model object
model.fitted <- model.example$copy()

# set up scenario options
start.date <- as.Date(c("2022-01-30", "2022-02-15"))
value <- c(1.5, 3)

sim <- ( 
  # chain two sudden changes in transmission successively
  sc(
    model = model.fitted, 
    opts = list(
    name = "sudden change in transmission",
    start.date = start.date[1],
    value = value[1]
  ))
  |> sc(
    model = model.fitted, 
    opts = list(
    name = "sudden change in transmission",
    start.date = start.date[2],
    value = value[2]
  ))
  # forecast from composed scenarios
  |> fcst_sc(model = model.fitted)
)

(sim
  # plot the forecast (using in-vignette utilities, not package functions!)
  |> prep_fcst(model=model.fitted, vars=c("Y"))
  |> plot_fcst()
  |> annotate_x(x = start.date, y.pos = c(0.65, 0.9),
  label = stringr::str_wrap(glue::glue("transmission increases suddenly by {value}x on {format(start.date, '%d %b')}"), width = 20))
  + labs(title = "Predicted incidence")
)
```